---
output: github_document
---

# earthtide #

The **earthtide** package simplifies the generation of earth tides and wave group harmonics for analysis in **R**.

[![Build Status](https://travis-ci.org/jkennel/earthtide.svg?branch=master)](https://travis-ci.org/jkennel/earthtide)
[![Coverage Status](https://img.shields.io/codecov/c/github/jkennel/earthtide/master.svg)](https://codecov.io/github/jkennel/earthtide?branch=master)

# Background #

**earthtide** is a port of the Fortran ETERNA 3.4 (Wenzel, 1996) predict and part of the analyze codes with the Kudryavtsev 2004 update.  The original Fortran code was rewritten in **R**, and **C++** using the great **Rcpp**, **RcppArmadillo**, and **RcppParallel**. The package is useful for generating synthetic earth tides using highly accurate tidal catalogs for prediction and regression applications in **R**. Attempts were made to ensure that results were consistent with the ETERNA 3.4 results, however, there is always the possibility that a bug was introduced in during the conversion and update.  For the most feature rich version and up-to-date version of ETERNA please see http://ggp.bkg.bund.de/eterna/ maintained by Klaus Schueller.

Wenzel, H.G. 1996: The nanogal software: Earth tide data processing package ETERNA 3.30. Bull. Inf. Marges Terrestres. 124, 9425-9439.

Kudryavtsev, S.M., 2004. Improved harmonic development of the Earth tide-generating potential. J. Geod. 77, 829–838.


# Installation #

```{r setup, echo = TRUE, eval = FALSE}
library(devtools)
install_github('jkennel/earthtide')
```


# Example # 

```{r example}
library(earthtide)

# One month of hourly data
tide <-  Earthtide$
  new(utc = as.POSIXct("2015-01-01", tz = "UTC") + 0:(24*31) * 3600,
      latitude = 52.3868,
      longitude = 9.7144,
      wave_groups = data.frame(start = 0.0, end = 6.0))$
  predict(method = "gravity", astro_update = 1)$
  lod_tide()$
  pole_tide()$
  tide()

```

```{r plot, echo = FALSE}
layout(matrix(1:3, ncol=1, nrow = 3))
par(mai = c(0.3, 0.9, 0.1, 0.1))
plot(gravity~datetime, tide, 
     ylab = expression('Gravity nm/s' ^ 2), 
     type = 'l',
     lwd  = 1.5,
     col  = '#5696BC',
     las  = 1)
plot(lod_tide~datetime, tide, 
     ylab = expression('LOD nm/s' ^ 2), 
     type = 'l',
     lwd  = 1.5,
     col  = '#5696BC',
     las  = 1)
plot(pole_tide~datetime, tide,
     ylab = expression('Pole nm/s' ^ 2), 
     type = 'l',
     lwd  = 1.5,
     col  = '#5696BC',
     las  = 1)
```

