---
output: github_document
---



```{r echo=FALSE, results="hide", message=FALSE}
library("badger")
```


```{r, echo = FALSE, results='asis'}

cat('# earthtide',
  '[![License: GPL v3](https://img.shields.io/badge/License-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)',
  badge_travis("jkennel/earthtide"),
  '[![Coverage Status](https://img.shields.io/codecov/c/github/jkennel/earthtide/master.svg)](https://codecov.io/github/jkennel/earthtide?branch=master)',
	badge_cran_release("earthtide", "green"),
	badge_cran_download("earthtide", "grand-total", "green"),
  badge_dependencies("earthtide")
)

```

The **earthtide** package simplifies the generation of earth tides and wave group harmonics for analysis in **R**.

## Background ##

**earthtide** is a port of the 'Fortran ETERNA 3.4' (Wenzel, 1996) predict and part of the analyze codes with the Kudryavtsev 2004 update.  The original 'Fortran' code was rewritten in **R**, and **C++** using the great **Rcpp**, **RcppArmadillo**, and **RcppParallel**. The package is useful for generating synthetic earth tides using highly accurate tidal catalogs for prediction and regression. Attempts were made to ensure that results were consistent with the 'ETERNA 3.4', however, there is always the possibility that a bug was introduced in during the conversion and update.  For the most feature rich and up-to-date version of 'ETERNA' please see http://ggp.bkg.bund.de/eterna/ maintained by Klaus Schueller.


Hartmann, T., Wenzel, H.-G., 1995. The HW95 tidal potential catalogue. Geophys. Res. Lett. 22, 3553–3556. \url(https://doi.org/10.1029/95GL03324)

Kudryavtsev, S.M., 2004. Improved harmonic development of the Earth tide-generating potential. J. Geod. 77, 829–838. \url(https://doi.org/10.1007/s00190-003-0361-2)

Wenzel, H.G. 1996: The nanogal software: Earth tide data processing package ETERNA 3.30. Bull. Inf. Marges Terrestres. 124, 9425-9439.  \url(http://www.eas.slu.edu/GGP/ETERNA34/MANUAL/ETERNA33.HTM)

## Installation ##

You can install the [CRAN](https://CRAN.R-project.org/package=earthtide) version with the following command:

```{r fromcran, echo = TRUE, eval = FALSE}
install.packages('earthtide')
```

The development version can be installed from github using the following commands:

```{r setup, echo = TRUE, eval = FALSE}
library(remotes)
remotes::install_github('jkennel/earthtide')
```

I will try to keep the IERS orientation datasets up-to-date in the development version (weekly update).

## Predict ## 

The following code computes the synthetic gravity, LOD tide, and pole tide and returns a `data.frame` of results. To see the list of possible arguments and methods of the Earthtide class use `?Earthtide`.

```{r example}
library(earthtide)

# One month of hourly data
tide <-  Earthtide$
  new(utc = as.POSIXct("2015-01-01", tz = "UTC") + 0:(24*31) * 3600,
      latitude = 52.3868,
      longitude = 9.7144,
      wave_groups = data.frame(start = 0.0, end = 6.0))$
  predict(method = "gravity", astro_update = 1)$    # compute gravity
  lod_tide()$                                       # LOD tide column
  pole_tide()$                                      # pole tide column
  tide()                                            # return result

```

```{r plot, echo = FALSE}
layout(matrix(1:3, ncol=1, nrow = 3))
par(mai = c(0.3, 0.9, 0.1, 0.1))
plot(gravity~datetime, tide, 
     ylab = expression('Gravity nm/s' ^ 2), 
     type = 'l',
     lwd  = 1.5,
     col  = '#5696BC',
     las  = 1)
plot(lod_tide~datetime, tide, 
     ylab = expression('LOD nm/s' ^ 2), 
     type = 'l',
     lwd  = 1.5,
     col  = '#5696BC',
     las  = 1)
plot(pole_tide~datetime, tide,
     ylab = expression('Pole nm/s' ^ 2), 
     type = 'l',
     lwd  = 1.5,
     col  = '#5696BC',
     las  = 1)
```

## Analyze ##

To generate curves for regression applications we can use the analyze method. For each wave group, normalized sin and cosine curves are generated.

```{r analyze, echo = TRUE}

wave_groups <- na.omit(eterna_wavegroups[eterna_wavegroups$time == '1 month', 
                                        c('start', 'end')])

tide_groups <-  Earthtide$
  new(utc = as.POSIXct("2015-01-01", tz = "UTC") + 0:(24*31) * 3600,
      latitude = 52.3868,
      longitude = 9.7144,
      wave_groups = wave_groups)$                                     
  analyze(method = 'gravity', astro_update = 1)$
  tide()

print(str(tide_groups))

```

