---
title: "Benchmarks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarks}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup}
library(earthtide)
library(bench)
```


# Catalog

Using a catalog with fewer waves will be faster. Here we compare ksm04 and hw95s.

```{r catalog}

tms <- as.POSIXct("1990-01-01", tz = "UTC") + 0:(1800)

wave_groups <- data.frame(start = 0, end = 8)

bench::mark(
  et_fun <- calc_earthtide(
    utc = tms,
    do_predict = TRUE,
    method = c("tidal_potential", "lod_tide", "pole_tide"),
    latitude = 52.3868,
    longitude = 9.7144,
    elevation = 110,
    gravity = 9.8127,
    cutoff = 1.0e-10,
    catalog = "ksm04",
    wave_groups = wave_groups
  ),
  et_fun_threads <- calc_earthtide(
    utc = tms,
    do_predict = TRUE,
    method = c("tidal_potential", "lod_tide", "pole_tide"),
    latitude = 52.3868,
    longitude = 9.7144,
    elevation = 110,
    gravity = 9.8127,
    cutoff = 1.0e-10,
    catalog = "hw95s",
    wave_groups = wave_groups
  ), check = FALSE
)

```


# cutoff parameter

Increasing the cutoff will decrease the number of wave groups and thus the speed
increases. Results will not be the same.

```{r cutoff}

tms <- as.POSIXct("1990-01-01", tz = "UTC") + 0:(1800)

wave_groups <- data.frame(start = 0, end = 8)

bench::mark(
  et_fun <- calc_earthtide(
    utc = tms,
    do_predict = TRUE,
    method = c("tidal_potential", "lod_tide", "pole_tide"),
    latitude = 52.3868,
    longitude = 9.7144,
    elevation = 110,
    gravity = 9.8127,
    cutoff = 1.0e-10,
    catalog = "ksm04",
    wave_groups = wave_groups
  ),
  et_fun_threads <- calc_earthtide(
    utc = tms,
    do_predict = TRUE,
    method = c("tidal_potential", "lod_tide", "pole_tide"),
    latitude = 52.3868,
    longitude = 9.7144,
    elevation = 110,
    gravity = 9.8127,
    cutoff = 1.0e-5,
    catalog = "ksm04",
    wave_groups = wave_groups
  ), check = FALSE
)

```



# astro_update parameter

Increasing the \code{astro_update} parameter leads to an approximation that may 
speed up computation. Results will not be exactly the same but can be very close 
as in the following example.

```{r astro_update}

tms <- as.POSIXct("1990-01-01", tz = "UTC") + 0:(1800)

wave_groups <- data.frame(start = 0, end = 8)

bench::mark(
  et_fun <- calc_earthtide(
    utc = tms,
    do_predict = TRUE,
    method = c("tidal_potential", "lod_tide", "pole_tide"),
    latitude = 52.3868,
    longitude = 9.7144,
    elevation = 110,
    gravity = 9.8127,
    cutoff = 1.0e-10,
    catalog = "ksm04",
    wave_groups = wave_groups
  ),
  et_fun_threads <- calc_earthtide(
    utc = tms,
    do_predict = TRUE,
    method = c("tidal_potential", "lod_tide", "pole_tide"),
    latitude = 52.3868,
    longitude = 9.7144,
    elevation = 110,
    gravity = 9.8127,
    cutoff = 1.0e-10,
    catalog = "ksm04",
    wave_groups = wave_groups,
    astro_update = 30L
  )
)


```

# n_thread parameter

Adjust the number of threads used for parallel computation. This should result
in equivalent values.

```{r threads}

tms <- as.POSIXct("1990-01-01", tz = "UTC") + 0:(1800)

wave_groups <- data.frame(start = 0, end = 8)

bench::mark(
  et_fun <- calc_earthtide(
    utc = tms,
    do_predict = TRUE,
    method = c("tidal_potential", "lod_tide", "pole_tide"),
    latitude = 52.3868,
    longitude = 9.7144,
    elevation = 110,
    gravity = 9.8127,
    cutoff = 1.0e-10,
    catalog = "ksm04",
    wave_groups = wave_groups
  ),
  et_fun_threads <- calc_earthtide(
    utc = tms,
    do_predict = TRUE,
    method = c("tidal_potential", "lod_tide", "pole_tide"),
    latitude = 52.3868,
    longitude = 9.7144,
    elevation = 110,
    gravity = 9.8127,
    cutoff = 1.0e-10,
    catalog = "ksm04",
    wave_groups = wave_groups,
    n_thread = 10L
  )
)


```

# Predict and interpolate

For one second output you can predict every minute and interpolate.  
Interpolation is done via \code{stat::spline}. This can also be done with longer
time intervals but the number of samples skipped may need to be adjusted. 
Results will not be the exactly the same but can be very close as in 
the following example.

```{r predictinterp}

tms <- as.POSIXct("1990-01-01", tz = "UTC") + 0:(1800)
tms_interp <- as.POSIXct("1990-01-01", tz = "UTC") + seq(0, 1800, 180)

wave_groups <- data.frame(start = 0, end = 8)

bench::mark(
  et_fun <- calc_earthtide(
    utc = tms,
    do_predict = TRUE,
    method = c("tidal_potential", "lod_tide", "pole_tide"),
    latitude = 52.3868,
    longitude = 9.7144,
    elevation = 110,
    gravity = 9.8127,
    cutoff = 1.0e-10,
    catalog = "ksm04",
    wave_groups = wave_groups
  ),
  et_fun_interp <- calc_earthtide(
    utc = tms_interp,
    do_predict = TRUE,
    method = c("tidal_potential", "lod_tide", "pole_tide"),
    latitude = 52.3868,
    longitude = 9.7144,
    elevation = 110,
    gravity = 9.8127,
    cutoff = 1.0e-10,
    catalog = "ksm04",
    wave_groups = wave_groups,
    utc_interp = tms
  )
)

  
```


# Combination of the above

We will use a larger dataset to compare a few approximation methods. In general, 
interpolation will give the best speed-up to accuracy if your time-steps are 
small (seconds).

```{r combination}

tms <- as.POSIXct("1990-01-01", tz = "UTC") + 0:(86400 * 2)
tms_interp <- as.POSIXct("1990-01-01", tz = "UTC") + seq(0, 86400 * 2, 180)

wave_groups <- data.frame(start = 0, end = 8)

bench::mark(
  et_fun_astro_threads <- calc_earthtide(
    utc = tms,
    do_predict = TRUE,
    method = c("tidal_potential", "lod_tide", "pole_tide"),
    latitude = 52.3868,
    longitude = 9.7144,
    elevation = 110,
    gravity = 9.8127,
    cutoff = 1.0e-10,
    catalog = "ksm04",
    wave_groups = wave_groups,
    astro_update = 60L,
    n_thread = 10L
  ),
  et_fun_interp_threads <- calc_earthtide(
    utc = tms_interp,
    do_predict = TRUE,
    method = c("tidal_potential", "lod_tide", "pole_tide"),
    latitude = 52.3868,
    longitude = 9.7144,
    elevation = 110,
    gravity = 9.8127,
    cutoff = 1.0e-10,
    catalog = "ksm04",
    wave_groups = wave_groups,
    utc_interp = tms,
    n_thread = 10L
  ),
)

  
```

