---
title: "Introduction to the *earthtide* package"
author: "Jonathan Kennel, Beth Parker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Code background 

**earthtide** is a port of the 'Fortran ETERNA 3.4' (Wenzel, 1996) predict and part of the analyze codes with the Kudryavtsev 2004 update.  The original Fortran code was rewritten in **R**, and **C++** using the great **Rcpp**, **RcppArmadillo**, and **RcppParallel**. The package is useful for generating synthetic earth tides using highly accurate tidal catalogs for prediction and regression applications in **R**. Attempts were made to ensure that results were consistent with the ETERNA 3.4 results, however, it is possible that a bug was introduced during the conversion and update.  For the most feature rich version and up-to-date version of ETERNA please see http://ggp.bkg.bund.de/eterna/ maintained by Klaus Schueller.


Hartmann, T., Wenzel, H.-G., 1995. The HW95 tidal potential catalogue. Geophys. Res. Lett. 22, 3553–3556. \url(https://doi.org/10.1029/95GL03324)

Kudryavtsev, S.M., 2004. Improved harmonic development of the Earth tide-generating potential. J. Geod. 77, 829–838. \url(https://doi.org/10.1007/s00190-003-0361-2)

Wenzel, H.G. 1996: The nanogal software: Earth tide data processing package ETERNA 3.30. Bull. Inf. Marges Terrestres. 124, 9425-9439. \url(http://www.eas.slu.edu/GGP/ETERNA34/MANUAL/ETERNA33.HTM)


## Setup

The first step to generate earthtides is to create the object with the prediction times, station, catalog and frequency details. The following are the most commonly used parameters:

- *utc* is a vector of POSIXct times for earth tide prediction.  This should typically be a regular time series (equally spaced) as the method is most efficient for this case, however, irregular time series are also accepted.
- *latitude* numeric of length one.  The latitude of the monitoring location.
- *longitude* numeric of length one.  The longitude of the monitoring location.
- *cutoff* numeric of length one. Cutoff determines how many waves to include in the analysis.  A lower number yields more precise results but at the cost of runtime. A value of 1e-10 includes all waves in the catalog and may be quite slow for long series. The default is 1e-6.
- *wave_groups* is a data.frame of of the frequency ranges in cycles per day. It requires a column named start and an column named end which give the ranges for each wave group.

See **?Earthtide** for the complete list of available arguments.

```{r newearthtide}
library(earthtide)

# One month of hourly data
tms <- as.POSIXct("2015-01-01", tz = "UTC") + 0:(24*31) * 3600

et <- Earthtide$new(
  utc = tms,
  latitude = 52.3868,
  longitude = 9.7144,
  wave_groups = data.frame(start = 0.0, end = 6.0))

```

## LOD (length of day) tide and pole tide

The Length of Day (LOD) and Pole tides are calculated in this initial step.

```{r lodpole, echo = FALSE, fig.width = 6.5, fig.height = 5}
tide <-et$lod_tide()$
  pole_tide()$
  tide()


layout(matrix(1:2, ncol=1, nrow = 2))
par(mai = c(0.4, 0.9, 0.1, 0.1))

# Plot the results

plot(lod_tide~datetime, tide,
     xlab = '',
     ylab = expression('LOD tide nm/s' ^ 2),
     type='l',
     lwd = 2,
     col = '#5696BC',
     las = 1)

plot(pole_tide~datetime, tide,
     xlab = '',
     ylab = expression('Pole tide nm/s' ^ 2),
     type='l',
     lwd = 2,
     col = '#5696BC',
     las = 1)

```

## Predict

The _**predict**_ method of the earthtide class computes the tidal response. There are two options for this function:

- *method* determines the which response is desired. Currently valid options are "gravity", "tidal_potential", "tidal_tilt", "vertical_displacement", "vertical_strain", "areal_strain", "volume_strain", and "ocean_tides". 
- *astro_update* determines how often phases are updated. For long time series (greater than 100,000 times) a value greater than 1 can speed up the results. In general, values greater than 1 will only be necessary for predicting at the minute or second scale or if you are predicting for many years.  There may be a slight loss of precision when the cutoff is small.  For small cutoff values use a small value for *astro_update* and we suggest testing to achieve the desired precision.
  

```{r predict}

et$predict(method = 'gravity', astro_update = 24)

```


```{r predictplot, fig.width = 6.5, fig.height = 3, fig.ext='png', echo = FALSE}
# Plot the results
par(mai = c(0.6, 0.9, 0.1, 0.1))
plot(gravity~datetime, et$tide(),
     ylab = expression('Gravity nm/s' ^ 2),
     xlab = '',
     type='l',
     lwd = 2,
     col = '#5696BC',
     xaxs = 'i',
     las = 1)
```


## Analyze

To generate normalized curves for use in regression equations we use the _**analyze**_ method instead of the _**predict**_ method.  All other arguments remain the same.  This will generate normalized sin and cos terms for each wavegroup.  In the example we plot the first 5 of a 22 wave groups set. These may be used in your preferred analyses.

```{r wg, fig.width = 6.5, fig.height = 8, fig.ext='png', echo = TRUE}
wave_groups <- na.omit(eterna_wavegroups[eterna_wavegroups$time == '1 month', 
                                        c('start', 'end')])

et <- Earthtide$new(utc = tms, 
                    latitude = 49.00937,
                    longitude = 8.40444,
                    elevation = 120,
                    cutoff = 1e-10,
                    catalog = 'hw95s',
                    wave_groups = wave_groups)

print(wave_groups[1:5,], row.names = FALSE)
```

```{r analyze}

et$analyze(method = 'gravity', astro_update = 1)

```

```{r analyzeplot, fig.width = 6.5, fig.height = 8, fig.ext='png', echo = FALSE}

tides <- et$tide()

layout(matrix(1:5, ncol=1, nrow = 5))
par(mai = c(0.3, 0.9, 0.1, 0.1))

for (i in seq(2, 11, 2)) {

  plot(tides[,1], tides[,i],
       ylab = expression('Gravity nm/s' ^ 2),
       xlab = '',
       type = 'l',
       lwd = 2,
       col = '#AAB6A2',
       las = 1)
  points(tides[,1], tides[,i+1],
         type = 'l',
         lwd = 2,
         col  = '#5696BC')

}


```


## Get the tidal values

The _**tide**_ method of the earthtide class computes returns a data.frame of tidal values. The following example in the chaining section shows how to calculate the gravity, LOD (length of day), and pole tides and return a data.frame of the values. 

## Chaining

It is also possible to chain operations together for more compact code.  

```{r r6chain, echo = TRUE}


tides <- Earthtide$
  new(utc = as.POSIXct("2015-01-01", tz = "UTC") + 0:(24*31) * 3600,
      latitude = 52.3868,
      longitude = 9.7144,
      wave_groups = data.frame(start = 0.0, end = 6.0))$
  predict(method = "gravity", astro_update = 1)$
  lod_tide()$
  pole_tide()$
  tide()
  

print(tides[1:5,], row.names = FALSE)

```


## TODO: 

- Speed enhancements
- Print methods and getter/setters for love numbers, catalog, station variables
- Optional wave group multipliers, wave group names
- Ocean loading
- Predicted Earth orientation data
